---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, echo = FALSE}
library(knitr)

opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```


# Dexter

Dexter is an R package for psychometric analysis of data from educational and psychological tests. Dexter typically works with project database files saved on disk. 

## Installation

```{r, eval=FALSE}
install.packages('dexter')
```

If you encounter a bug, please post a minimal reproducible example on [github](https://github.com/jessekps/dexter/issues). We post news and examples on a [blog](http://dexterities.netlify.com), it's also the place for general questions.

## Example


```{r, results='hide', message=FALSE}
library(dexter)
# start a project and fill it with data
# verbAggrRules and verbAggrData are example datasets provided with dexter

db = start_new_project(verbAggrRules, "verbAggression.db")
add_booklet(db, verbAggrData, booklet_id = "verb_agg")

# Classical test theory
tia = tia_tables(db)

tia$testStats
```




```{r, echo=FALSE}
kable(tia$testStats, digits=3,table.attr='class="myTable"')
```

```{r, eval=FALSE}
head(tia$itemStats)
```

```{r, echo=FALSE}
kable(head(tia$itemStats), digits=3)
```


```{r, results='hide'}
# IRT, extended nominal response model
f = fit_enorm(db)

head(coef(f))
```

```{r, echo=FALSE}
kable(head(coef(f)))
```

```{r, results='hide'}
# ability estimates per person
abl = ability(db, parms = f)
head(abl)
```

```{r, echo=FALSE}
kable(head(abl))
```

```{r, results='hide'}
# ability estimates without item S1DoScold
abl2 = ability(db, parms = f, item_id != "S1DoScold")

# plausible values

pv = plausible_values(db, parms = f, nPV = 5)
head(pv)
```

```{r, echo=FALSE}
kable(head(pv))
```


```{r, include=FALSE}
close_project(db)
```
## Contributing

Contributions are welcome but please check with us first about what you would like to contribute.


